A semantics of core AUTOSAR

### 2016-08-10 paper/paper.tex
Major changes to server invocation and return semantics.

The process terms of our calculus are as follows:
\begin{itemize}
\item $\Trunnable{\Tr{\V{R}}{\V{I}}}{\V{T}}{\Act}{\V{N}}$  \newline
Dynamic state for runnable \V{R} of component \V{I}, indicating \V{N} current instances, \V{T} seconds left until next possible instantiation, and activation state $\Act$.
\item $\Trinst{\Tr{\V{R}}{\V{I}}}{\V{C}}{\Xs}{\Code}$ \newline
An instance of runnable \V{R} of component \V{I}, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client \V{C}.
\item $\Texcl{\Tx{\V{X}}{\V{I}}}{\V{U}}$ \newline
Exclusive area \V{X} of component \V{I}, with boolean occupation state \V{U}.
\item $\Tirv{\Ts{\V{S}}{\V{I}}}{\V{V}}$ \newline
Inter-runnable variable \V{S} of component \V{I}, currently holding value \V{V}.
\item $\Tdelem{\Tep{\V{E}}{\V{P}}{\V{I}}}{\V{U}}{\V{V}}$ \newline
Non-queued sender-receiver data element \V{E} in port \V{P} of component \V{I}, currently holding value \V{V} with boolean update status flag \V{U}.
\item $\Tqelem{\Tep{\V{E}}{\V{P}}{\V{I}}}{\V{N}}{\Vs}$ \newline
Queued sender-receiver data element \V{E} in port \V{P} of component \V{I} with maximum capacity \V{N}, currently holding value sequence \Vs.
\item $\Topres{\Top{\V{O}}{\V{P}}{\V{I}}}{\Vs}$ \newline
Client-server operation \V{O} in port \V{P} of component \V{I}, currently holding result value sequence \Vs.
\item $\Ttimer{\Tr{\V{R}}{\V{I}}}{\V{T_p}}{\V{T}}$ \newline
Periodic timer for runnable \V{R} of component \V{I}, with a period of \V{T_p} and \V{T} seconds left of its current cycle.
\end{itemize}

In the untyped setting of this report, a value \V{V} can be any data item computed and communicated by an AUTOSAR system, including the unit value $\Tvoid$ and the error codes that might be returned from RTE calls. Meta-variables \V{N}, \V{U} and \V{T} range over natural numbers, boolean values and (floating point) time values, respectively. The client \V{C} of a runnable instance is the address of the invoking port operation (if the runnable was invoked by a client-server call), or $\Tvoid$ otherwise. An activation state $\Act$ either toggles between $\Tidle$ and $\Tpending$, or it has the special shape $\Tserving{\Cs}{\Vs}$ (in the case of a runnable triggered by client-server calls). Formally,
\[
\begin{array}{rcll}
  \V{V}    & \in & \text{Values} \\
  \V{N}    & \in & \text{Natural numbers} \\
  \V{U}    & \in & \text{Boolean values} \\
  \V{T}    & \in & \text{Time values} \\
  \V{C}    & \in & \text{Clients} \\
  \Act     & \in & \text{Activation states} \\ \\
  \V{C}    & ::= & \Top{\V{O}}{\V{P}}{\V{I}} \sep \Tvoid \\
  \Act     & ::= & \Tidle \sep \Tpending \sep \Tserving{\Cs}{\Vs} \\
\end{array}
\]
We write $\Vs$ for a sequence of values \V{V}, $\Cs$ for a sequence of clients \V{C}, etc. Sequences may be empty, which is written $\epsilon$. By $\Tlength{\Vs}$ we mean the length of sequence $\Vs$. Left (right) concatenation of an element with a sequence is written $\Tcons{\V{V}}{\V{Vs}}$ ($\Tsnoc{\V{Vs}}{\V{V}}$).

@@

To this end, we have chosen to ignore all internal computations and just capture the observable effects of runnable execution as a pure sequence of RTE (VFB) calls. And because the result from an RTE call may in general affect subsequent runnable behavior, we use a continuation-passing style where every $\Code$ term except the final function return actually contains the sequence that follows it as a continuation function $\Cont$.
%
\[
\begin{array}{lll}
  \Code & \in & \text{Code}								\\
  \Cont & \in & \text{Values} \rightarrow \text{Code}   \\ \\
  \Code & ::= & \Enter{\V{X}}{\Cont}                \\
        & |   & \Exit{\V{X}}{\Cont}                 \\
        & |   & \IrvWrite{\V{S}}{\V{V}}{\Cont}          \\
        & |   & \IrvRead{\V{S}}{\Cont}              \\
        & |   & \Send{\Te{\V{E}}{\V{P}}}{\V{V}}{\Cont}      \\
        & |   & \Receive{\Te{\V{E}}{\V{P}}}{\Cont}      \\
        & |   & \Write{\Te{\V{E}}{\V{P}}}{\V{V}}{\Cont}     \\
        & |   & \Read{\Te{\V{E}}{\V{P}}}{\Cont}         \\
        & |   & \Invalidate{\Te{\V{E}}{\V{P}}}{\Cont}   \\
        & |   & \IsUpdated{\Te{\V{E}}{\V{P}}}{\Cont}    \\
        & |   & \Call{\To{\V{O}}{\V{P}}}{\V{V}}{\Cont}      \\
        & |   & \Result{\To{\V{O}}{\V{P}}}{\Cont}       \\
        & |   & \Return{\V{V}}                      \\
\end{array}
\]

In contrast to $\Code$ sequences, process terms are completely unordered. This is expressed in the standard process calculus style, using an associative operator $\Tpar{}{}$ that allows arbitrary sets of primitive processes to be composed in parallel. This operator also has a left and right unit in the form of process $\Tzero$, which stands for the empty, or terminated, process. The complete grammar for our process terms thus looks as follows:

\[
\begin{array}{rcll}
  \V{P},\V{Q}   
  		& ::= & \Trunnable{\V{A}}{\V{T}}{\Act}{\V{N}}    \\
        & |   & \Trinst{\V{A}}{\V{C}}{\Xs}{\Code}       \\
        & |   & \Texcl{\V{A}}{\V{U}}                    \\
        & |   & \Tirv{\V{A}}{\V{V}}                     \\
        & |   & \Tdelem{\V{A}}{\V{U}}{\V{V}}            \\
        & |   & \Tqelem{\V{A}}{\V{N}}{\Vs}              \\
        & |   & \Topres{\V{A}}{\Vs}                     \\
        & |   & \Ttimer{\V{A}}{\V{T_p}}{\V{T}}          \\
        & |   & \Tpar{\V{P}}{\V{Q}}                     \\
        & |	  & \Tzero									\\
\end{array}
\]

@@

Our semantic approach defines the meaning of an AUTOSAR system as the possible chains of state changes that can be applied to the system's initial state; or equivalently, as the set of \emph{traces} that can be generated from the initial process term \V{P_0}. A trace is a possibly infinite sequence of transitions
\[
\begin{array}{c}
  \V{P_0} \red{\V{L_1}} \V{P_1} \red{\V{L_2}} \V{P_2} \red{\V{L_3}} \cdots
\end{array}
\]
where each step
\[
\begin{array}{c}
  \V{P_{i-1}} \red{\V{L_i}} \V{P_i}
\end{array}
\]
states that the system state captured as process term \V{P_{i-1}} can evolve into state \V{P_i} by means of a single transition labelled \V{L_i}.

The label \V{L} of a transition can essentially be of two kinds, indicating that the affected process either "says" or "hears" something during the transition. Such a label always consists of an address \V{A} paired with some additional detail \V{D}. There is also a special form of "hearing" that denotes letting time pass and which does not carry any address. The different labels are captured in the following grammar. 
\[
\begin{array}{rcll}
  \V{L}     & ::= & \V{A}!\V{D} 	& \hspace{4em}\text{Say  \V{A} and \V{D}} \\
            & |   & \V{A}?\V{D} 	& \hspace{4em}\text{Hear \V{A} and \V{D}} \\
            & |   & \Tdelta{\V{T}}	& \hspace{4em}\text{Let \V{T} seconds pass} \\
\end{array}
\]
The possible forms of the extra label information \V{D} will be introduced as the different transition rules are defined.

Two processes \V{P} and \V{Q} can make transitions in parallel if they agree on what is being said or heard. Just as the saying/hearing intuition suggests, at most one of the processes can have the saying role in such an agreement. The following set of inference rules capture this intuition formally.
{
\renewcommand{\Prule}[2]{#1 \quad \Pif\; #2\\}
\renewcommand{\Tstep}[3]{#1 \red{#2} #3}
\renewcommand\Pcomma{\;\text{and}\;}

@@

The initial process \V{P_0} is determined fully by the AUTOSAR model under study, as the parallel composition of the following terms:
\begin{enumerate}

\item For each runnable \V{R} of each component \V{I}, a term
\[
\begin{array}{c}
  \Trunnable{\Tr{\V{R}}{\V{I}}}{0}{\Act}{0}
\end{array}
\]
where $\Act$ is $\Tserving{\epsilon}{\epsilon}$ if \V{R} is triggered by an \emph{OperationInvokedEvent}; else $\Act$ is $\Tpending$ if \V{R} is triggered by an \emph{InitEvent}; otherwise $\Act$ is $\Tidle$.
% SWS_Rte_03524 (SWS_Rte_03526?) forbids mixing OperationInvokedEvents with other events for the same runnable

\item For each exclusive area \V{X} of each atomic component \V{I}, a term
\[
\begin{array}{c}
  \Texcl{\Tx{\V{X}}{\V{I}}}{\Tfalse}
\end{array}
\]

\item For each inter-runnable variable \V{S} of each atomic component \V{I}, a term
\[
\begin{array}{c}
  \Tirv{\Ts{\V{S}}{\V{I}}}{\V{V}}
\end{array}
\]
where \V{V} is the \emph{initValue} attributed to \V{S} if it exists, otherwise \V{V} is $\Tvoid$.

\item For each data element \V{E} of each required port \V{P} of each atomic component \V{I}, a term
\[
\begin{array}{c}
  \Tqelem{\Tep{\V{E}}{\V{P}}{\V{I}}}{\V{N}}{\epsilon}
\end{array}
\]
if the \emph{swImplPolicy} attribute of element \V{E} is \emph{queued} (with capacity \V{N}); otherwise, a term
\[
\begin{array}{c}
  \Tdelem{\Tep{\V{E}}{\V{P}}{\V{I}}}{\Tfalse}{\TneverReceived}
\end{array}
\]

\item For each operation \V{O} of each required port \V{P} of each atomic component \V{I}, a term
\[
\begin{array}{c}
  \Topres{\Tep{\V{O}}{\V{P}}{\V{I}}}{\epsilon}
\end{array}
\]

\item For each \emph{TimingEvent} triggering each runnable \V{R} of each atomic component \V{I}, a term
\[
\begin{array}{c}
  \Ttimer{\Tr{\V{R}}{\V{I}}}{\V{T_p}}{0}
\end{array}
\]
where \V{T_p} is the period of the event.
%TODO: This means that all timed runnables want to act _immediately_ (at the same time - with no init delay).

@@

An inter-runnable variable (irv) represents state that is local to a particular component and accessible by all runnables of that component. Axioms (5) and (6) show the \emph{say} transitions taken by a runnable instance wishing to read and write an irv $s$. In both cases the embedded continuation $k$ is applied to the command result. The corresponding \emph{hear} transitions of an irv appear as rules (7) and (8). Thanks to inference rule (1), a runnable instance and an irv can now engage in a parallel transition, for example expressing the writing of shared data.
$$
\begin{array}{c}
	\Tpar {\Trinst {i.r} c \Xs {\TrteIrvWrite s v k}}
	 	  {\Tirv {i.s} \_} \vspace{0.6ex} \\ 
	\red {\say {i.s} {\Tirvw v}} \\
	\Tpar {\Trinst {i.r} c \Xs {\Tap k \Tvoid}}
		  {\Tirv {i.s} v}
\end{array}
$$
The corresponding read transition looks as follows:
$$
\begin{array}{c}
	\Tpar {\Trinst {i.r} c \Xs {\TrteIrvRead s k}}
	 	  {\Tirv {i.s} v} \vspace{0.6ex} \\ 
	\red{\say {i.s} {\Tirvr v}} \\
	\Tpar{\Trinst {i.r} c \Xs {\Tap k v}}
		 {\Tirv {i.s} v}
\end{array}
$$
Section~\ref{sect:Ignore} will show how examples like these can be extended to the case when both participating processes are embedded in arbitrarily large parallel process compositions.

Note that in the last example above, the read value $v$ flows from the irv process to the runnable instance, even though axiom (7) states that an irv actually \emph{hears} the transition payload $\Tirvr{v}$. This apparent paradox is simply a reminder that the saying/hearing distinction of our calculus, and the dataflow patterns it may define, are entirely separate notions.

@@

An exclusive area is the AUTOSAR equivalent of a binary semaphore, that can be acquired and released in an atomic fashion by competing runnable instances. Axiom (9) states that a runnable instance wishing to enter exclusive area $x$ may successfully proceed to its continuation $k$ by broadcasting $\say{\Tx x i}{\Tent}$. Axiom (11) expresses that exclusive area $\Tx x i$ accepts hearing an ${\Tent}$ payload if it is currently not taken; i.e., carries the boolean state $\Tfalse$. This makes the following parallel transition possible:
$$
\begin{array}{c}
	\Trinst {i.r} c \Xs {\TrteEnter x k}
	\Opar
	\Texcl {i.x} \Tfalse \vspace{0.6ex} \\ 
	\red{\say {i.x} \Tent} \\
	\Trinst{i.r} c {\Tcons x \Xs} {\Tap k \Tvoid}
	\Opar
	\Texcl {i.x} \Ttrue
\end{array}
$$
Exiting from an exclusive area is simply the reverse, as defined in axioms (10) and (12):
$$
\begin{array}{c}
	\Trinst {i.r} c {\Tcons x \Xs} {\TrteExit x k}
	\Opar
	\Texcl {i.x} \Ttrue \vspace{0.6ex} \\ 
	\red{\say {i.x} \Tex} \\
	\Trinst{i.r} c \Xs {\Tap k \Tvoid}
	\Opar
	\Texcl {i.x} \Tfalse
\end{array}
$$
However, because axiom (11) is only defined for exclusive areas in the not taken state, there is no way to derive any parallel transition of the following form:
$$
\begin{array}{c}
	p \Opar {\Texcl {i.x} \Ttrue} \red{\say{i.x}{\Tent}} \ldots \\
\end{array}
$$
This means that any runnable instance in $p$ wishing to enter $x$ is effectively blocked from making progress until some other process in $p$ choses to exit $x$. In the same manner, a process wishing to exit an exclusive area not taken is also effectively blocked. The conditions on exiting are actually stronger than those that govern entering, since axiom (10) is only applicable to a runnable instance for which the exited area $x$ is at the top of its stack of acquired exclusive areas (as expressed by the sequence pattern $\Tcons x \Xs$). An exclusive area can thus only be exited by the process that entered it, and only in the reverse order of entering. Section~\ref{sect:DiscExcl} will bring up some alternatives to this semantics, and also discuss the interpretation of blocked processes in more detail.

@@

Axioms (13) and (14) define unbuffered interprocess communication from a runnable instance point of view. These transitions are just syntactic variants of shared variable reading and writing (axioms (5) and (6)). A data element ({\bf delem}) process term contains two pieces of state in addition to its address: a communicated data value and a boolean flag for keeping track of unread writes. Axioms (15) and (16) capture reads and writes to the data value and also set the update flag accordingly. The side-confition in axiom (16) makes the rule apply only if there is a static connection from the data element being written and the address of the matched term (recall that data element storage is associated with the receiving side of a connection). Because AUTOSAR allows sender-receiver communication patterns such that $\Tconnect {i.p.e} a$ and $\Tconnect {i.p.e} b$ are simultaneously true, parallel data element updates like the following are possible:
$$
\begin{array}{c}
	\Trinst {i.r} c \Xs {\TrteWrite {p.e} v k} \,\Opar  \\
	{\Tdelem a \_ \_} \Opar {\Tdelem b \_ \_} \vspace{0.6ex} \\
	\red{\say{i.p.e}{\Twr v}} \\
	\Trinst {i.r} c \Xs {\Tap k \Tvoid} \,\Opar \\
	{\Tdelem a \Ttrue v} \Opar {\Tdelem b \Ttrue v}
\end{array}
$$
As an aside, note that the associativity of the parallel composition operator is made entirely irrelevant by the symmetric shape of transition rules (1) and (2).

Axiom (17) provides another means to intepret a data element write. It expresses that a runnable set up to trigger on data reception on required data element $\Tep e p i$ should be marked as {\bf pending} whenever a write occurs on an element connected to $\Tep e p i$. An example may be as follows (assuming $\Tconnect {i.p.e} a$ and $\TdataReceivedEvent b a$:
$$
\begin{array}{c}
	\Trinst {i.r} c \Xs {\TrteWrite {p.e} v k} \,\Opar  \\
	{\Tdelem a \_ \_} \Opar {\Trunnable b t \_ n} \vspace{0.6ex} \\
	\red{\say{i.p.e}{\Twr v}} \\
	\Trinst {i.r} c \Xs {\Tap k \Tvoid} \,\Opar \\
	{\Tdelem a \Ttrue v} \Opar {\Trunnable b t \Tpending n}
\end{array}
$$
Transitions that actually create an instance of a pending runnable will be introduced in Section~\ref{sect:SpawnTerm}.

@@

Buffered interprocess communication (axioms (22) - (30)) is essentially a minor variant of the unbuffered mechanism, where the single-value store of a {\bf delem} term is replaced by a {\bf qelem} process holding a sequence of values. The buffer is consumed from the left (axiom (25)) and extended to the right (axiom (27)). When the buffer is empty, the special value (i.e., the error code) {\bf nodata} is returned to receiving runnable instances (axiom (26)). When the buffer is at its maximum capacity and cannot hold more values, the provided value is simply dropped (axiom (28)).

However, the AUTOSAR specification requires

@@

>>>>

The process terms of our calculus are as follows:
\begin{itemize}
\item $\Trunnable {\Tr r i} t \Act n$  \newline
Dynamic state for runnable $r$ of component $i$, indicating $n$ current instances, $t$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\Trinst {\Tr r i} c \Xs \Code$ \newline
An instance of runnable $r$ of component $i$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $c$.
\item $\Texcl {\Tx x i} u$ \newline
Exclusive area $x$ of component $i$, with boolean occupation state $u$.
\item $\Tirv {\Ts s i} v$ \newline
Inter-runnable variable $s$ of component $i$, currently holding value $v$.
\item $\Tdelem {\Tep e p i} u v$ \newline
Non-queued sender-receiver data element $e$ in port $p$ of component $i$, currently holding value \V{V} with boolean update status flag $u$.
\item $\Tqelem {\Tep e p i} n \Vs$ \newline
Queued sender-receiver data element $e$ in port $p$ of component $i$ with maximum capacity $n$, currently holding value sequence $\Vs$.
\item $\Toper {\Top o p i} m \Srv$ \newline
Client-server operation $o$ in port $p$ of component $i$, currently holding sequence number $m$ and call state $\Srv$.
\item $\Ttimer {\Tr r i} {t_p} t$ \newline
Periodic timer for runnable $r$ of component $i$, with a period of $t_p$ and $t$ seconds left of its current cycle.
\end{itemize}

In the untyped setting of this report, a value $v$ can be any data item computed and communicated by an AUTOSAR system, including the unit value $\Tvoid$ and the error codes that might be returned from RTE calls. Meta-variables $n$ and $m$ stand for natural numbers, while $u$ and $t$ range over boolean values and (floating point) time values, respectively. The client $c$ of a runnable instance is a tuple holding the address, sequence number and argument value of the current invocation (if the runnable was invoked by a client-server call), or is $\Tvoid$ otherwise. An activation state $\Act$ either toggles between $\Tidle$ and $\Tpending$, or is a sequence of client tuples (in the case of a runnable triggered by client-server calls). A call state $\Srv$ alternates between indicating an ongoing call with a timeout and a finished call with a result value.
Formally,
\[
\begin{array}{rcll}
  v        & \in & \text{Values} \\
  n,m      & \in & \text{Natural numbers} \\
  u        & \in & \text{Boolean values} \\
  t        & \in & \text{Time values} \\
  c        & \in & \text{Clients} \\
  \Act     & \in & \text{Activation states} \\
  \Srv	   & \in & \text{Call states} \\ \\
  c        & ::= & \Tclient {\Top o p i} m v \sep \Tvoid \\
  \Act     & ::= & \Tidle \sep \Tpending \sep \Cs \\
  \Srv	   & ::= & \Tcalling t \sep \Tdone v
\end{array}
\]
We write $\Vs$ for a sequence of values \V{V}, $\Cs$ for a sequence of clients \V{C}, etc. Sequences may be empty, which is written $\epsilon$. By $\Tlength{\Vs}$ we mean the length of sequence $\Vs$. Left (right) concatenation of an element with a sequence is written $\Tcons{\V{V}}{\V{Vs}}$ ($\Tsnoc{\V{Vs}}{\V{V}}$).

@@

To this end, we have chosen to ignore all internal computations and just capture the observable effects of runnable execution as a pure sequence of RTE (VFB) calls. And because the result from an RTE call may in general affect subsequent runnable behavior, we use a continuation-passing style where every $\Code$ term except the final function return literally the sequence that follows it, as a continuation function $\Cont$.
%
\[
\begin{array}{lll}
  \Code & \in & \text{Code}								\\
  \Cont & \in & \text{Values} \rightarrow \text{Code}   \\ \\
  \Code & ::= & \Enter x \Cont               \\
        & |   & \Exit x \Cont                \\
        & |   & \IrvWrite s v \Cont          \\
        & |   & \IrvRead s \Cont             \\
        & |   & \Send {\Te e p} v \Cont      \\
        & |   & \Receive {\Te e p} \Cont     \\
        & |   & \Write {\Te e p} v \Cont     \\
        & |   & \Read {\Te e p} \Cont        \\
        & |   & \Invalidate {\Te e p} \Cont  \\
        & |   & \IsUpdated {\Te e p} \Cont   \\
        & |   & \Call {\To o p} v \Cont      \\
        & |   & \Result {\To o p} \Cont      \\
        & |   & \Return v                    \\
\end{array}
\]

In contrast to $\Code$ sequences, process terms are completely unordered. This is expressed in the standard process calculus style using an associative operator $\Tpar{}{}$ that allows arbitrary sets of primitive processes to be composed in parallel. This operator also has a left and right unit in the form of process $\Tzero$, which stands for the empty, or terminated, process. The complete grammar for our process terms thus looks as follows:

\[
\begin{array}{rcll}
  \V{P},\V{Q}   
  		& ::= & \Trunnable a t {\Act} n	                \\
        & |   & \Trinst a c {\Xs} {\Code}               \\
        & |   & \Texcl a u                              \\
        & |   & \Tirv a v                               \\
        & |   & \Tdelem a u v                           \\
        & |   & \Tqelem a n {\Vs}                       \\
        & |   & \Toper a m \Srv                         \\
        & |   & \Ttimer a {\V{T_p}} t                   \\
        & |   & \Tpar p q                               \\
        & |	  & \Tzero									\\
\end{array}
\]

@@

Our semantic approach defines the meaning of an AUTOSAR system as the possible chains of state changes that can be applied to the system's initial state; or equivalently, as the set of \emph{traces} that can be generated from the initial process term \V{P_0}. A trace is a possibly infinite sequence of transitions
\[
\begin{array}{c}
  {p_0} \red{l_1} {p_1} \red{l_2} {p_2} \red{l_3} \cdots
\end{array}
\]
where each step
\[
\begin{array}{c}
  {p_{i-1}} \red{l_i} {p_i}
\end{array}
\]
states that the system state captured as process term $p_{i-1}$ can evolve into state $p_i$ by means of a single transition labelled $l_i$.

The label $l$ of a transition can essentially be of two kinds, indicating that the affected process either "says" or "hears" something during the transition. Such a label always consists of an address $a$ paired with some additional detail $d$. There is also a special form of "hearing" that denotes letting time pass and which does not carry any address. The different labels are captured in the following grammar. 
\[
\begin{array}{rcll}
  l         & ::= & a!d 	    & \hspace{4em}\text{Say  \V{A} and \V{D}} \\
            & |   & a?d 	    & \hspace{4em}\text{Hear \V{A} and \V{D}} \\
            & |   & \Tdelta t	& \hspace{4em}\text{Let \V{T} seconds pass} \\
\end{array}
\]
The possible forms of the extra label information \V{D} will be introduced as the different transition rules are defined.

Two processes $p$ and $q$ can make transitions in parallel if they agree on what is being said or heard. Just as the saying/hearing intuition suggests, at most one of the processes can have the saying role in such an agreement. The following set of inference rules capture this intuition formally.
{
\renewcommand{\Prule}[2]{#1 \quad \Pif\; #2\\}
\renewcommand{\Tstep}[3]{#1 \red{#2} #3}
\renewcommand\Pcomma{\;\text{and}\;}

@@

The initial process $p_0$ is determined fully by the AUTOSAR model under study, as the parallel composition of the following terms:
\begin{enumerate}

\item For each runnable $r$ of each component $i$, a term
\[
\begin{array}{c}
  \Trunnable {\Tr r i} 0 \Act 0
\end{array}
\]
where $\Act$ is $\epsilon$ if $r$ is triggered by an \emph{OperationInvokedEvent}; else $\Act$ is $\Tpending$ if $r$ is triggered by an \emph{InitEvent}; otherwise $\Act$ is $\Tidle$.
% SWS_Rte_03524 (SWS_Rte_03526?) forbids mixing OperationInvokedEvents with other events for the same runnable

\item For each exclusive area $x$ of each atomic component $i$, a term
\[
\begin{array}{c}
  \Texcl {\Tx x i} \Tfalse
\end{array}
\]

\item For each inter-runnable variable $s$ of each atomic component $i$, a term
\[
\begin{array}{c}
  \Tirv {\Ts s i} v
\end{array}
\]
where $v$ is the \emph{initValue} attributed to $s$ if it exists, otherwise $v$ is $\Tvoid$.

\item For each data element $e$ of each required port $p$ of each atomic component $i$, a term
\[
\begin{array}{c}
  \Tqelem {\Tep e p i} n \epsilon
\end{array}
\]
if the \emph{swImplPolicy} attribute of element $e$ is \emph{queued} (with capacity $n$); otherwise, a term
\[
\begin{array}{c}
  \Tdelem {\Tep e p i} \Tfalse \TneverReceived
\end{array}
\]

\item For each operation $o$ of each required port $p$ of each atomic component $i$, a term
\[
\begin{array}{c}
  \Toper{\Top o p i} 0 {\Tdone \Tvoid}
\end{array}
\]

\item For each \emph{TimingEvent} of each runnable $r$ of each atomic component $i$, a term
\[
\begin{array}{c}
  \Ttimer {\Tr r i } {t_p} 0
\end{array}
\]
where $\V{T_p}$ is the period of the event.
%TODO: This means that all timed runnables want to act _immediately_ (at the same time - with no init delay).

@@

An inter-runnable variable (irv) represents state that is local to a particular component and accessible by all runnables of that component. Axioms (5) and (7) show the \emph{say} transitions taken by a runnable instance wishing to read and write an irv $s$. In both cases the embedded continuation $k$ is applied to the command result. The corresponding \emph{hear} transitions of an irv appear as rules (6) and (8). Thanks to inference rule (1), a runnable instance and an irv can now engage in a parallel transition, for example expressing the writing of shared data.
$$
\begin{array}{c}
	\Tpar {\Trinst {i.r} c \Xs {\TrteIrvWrite s v k}}
	 	  {\Tirv {i.s} \_} \vspace{0.6ex} \\ 
	\red {\say {i.s} {\Tirvw v}} \\
	\Tpar {\Trinst {i.r} c \Xs {\Tap k \Tvoid}}
		  {\Tirv {i.s} v}
\end{array}
$$
The corresponding read transition looks as follows:
$$
\begin{array}{c}
	\Tpar {\Trinst {i.r} c \Xs {\TrteIrvRead s k}}
	 	  {\Tirv {i.s} v} \vspace{0.6ex} \\ 
	\red {\say {i.s} {\Tirvr v}} \\
	\Tpar {\Trinst {i.r} c \Xs {\Tap k v}}
		  {\Tirv {i.s} v}
\end{array}
$$
Section~\ref{sect:Ignore} will show how examples like these can be extended to the case when both participating processes are embedded in arbitrarily large parallel process compositions.

Note that in the last example above, the read value $v$ flows from the irv process to the runnable instance, even though axiom (6) states that an irv actually \emph{hears} the transition payload $\Tirvr{v}$. This apparent paradox is simply a reminder that the saying/hearing distinction of our calculus is entirely separate from the dataflow patterns it defines.

@@

An exclusive area is the AUTOSAR equivalent of a binary semaphore, that can be acquired and released in an atomic fashion by competing runnable instances. Axiom (9) states that a runnable instance wishing to enter exclusive area $x$ may successfully proceed to its continuation $k$ by broadcasting $\say {i.x}{\Tent}$. Axiom (10) expresses that exclusive area $i.x$ accepts hearing an ${\Tent}$ payload if it is currently not taken; i.e., carries the boolean state $\Tfalse$. This makes the following parallel transition possible:
$$
\begin{array}{c}
	\Trinst {i.r} c \Xs {\TrteEnter x k}
	\Opar
	\Texcl {i.x} \Tfalse \vspace{0.6ex} \\ 
	\red {\say {i.x} \Tent} \\
	\Trinst {i.r} c {\Tcons x \Xs} {\Tap k \Tvoid}
	\Opar
	\Texcl {i.x} \Ttrue
\end{array}
$$
Exiting from an exclusive area is simply the reverse, as defined in axioms (11) and (12):
$$
\begin{array}{c}
	\Trinst {i.r} c {\Tcons x \Xs} {\TrteExit x k}
	\Opar
	\Texcl {i.x} \Ttrue \vspace{0.6ex} \\ 
	\red {\say {i.x} \Tex} \\
	\Trinst {i.r} c \Xs {\Tap k \Tvoid}
	\Opar
	\Texcl {i.x} \Tfalse
\end{array}
$$

Because axiom (10) is only defined for exclusive areas in the not taken state, there is no way to derive any parallel transition of the following form:
$$
\begin{array}{c}
	p \Opar {\Texcl {i.x} \Ttrue} \red{\say{i.x}{\Tent}} \ldots \\
\end{array}
$$
This means that any runnable instance in $p$ wishing to enter $x$ is effectively blocked from making progress until some other process in $p$ choses to exit $x$. In the same manner, a process wishing to exit an exclusive area not taken is also effectively blocked. The conditions on exiting are actually stronger than those that govern entering, since axiom (11) is only applicable to a runnable instance for which the exited area $x$ is at the top of its stack of acquired exclusive areas (as expressed by the sequence pattern $\Tcons x \Xs$). An exclusive area can thus only be exited by the process that entered it, and only in the reverse order of entering. Section~\ref{sect:DiscExcl} will bring up some alternatives to this semantics, and also discuss the interpretation of blocked processes in more detail.

@@

Axioms (13) and (15) define unbuffered interprocess communication from a runnable instance point of view. These transitions are just syntactic variants of shared variable reading and writing (axioms (5) and (7)). A data element ({\bf delem}) process term contains two pieces of state in addition to its address: a communicated data value and a boolean flag for keeping track of unread writes. Axioms (14) and (16) capture reads and writes to the data value and also set the update flag accordingly. The side-confition in axiom (16) makes the rule apply only if there is a static connection from the data element being written and the address of the matched term (recall that data element storage is associated with the receiving side of a connection). Because AUTOSAR allows sender-receiver communication patterns such that $\Tconnect {i.p.e} a$ and $\Tconnect {i.p.e} b$ are simultaneously true, parallel data element updates like the following are possible:
$$
\begin{array}{c}
	\Trinst {i.r} c \Xs {\TrteWrite {p.e} v k} \,\Opar  \\
	{\Tdelem a \_ \_} \Opar {\Tdelem b \_ \_} \vspace{0.6ex} \\
	\red{\say{i.p.e}{\Twr v}} \\
	\Trinst {i.r} c \Xs {\Tap k \Tvoid} \,\Opar \\
	{\Tdelem a \Ttrue v} \Opar {\Tdelem b \Ttrue v}
\end{array}
$$
As an aside, note that the associativity of the parallel composition operator is made entirely irrelevant by the symmetric shape of transition rules (1) and (2).

Axiom (17) provides another means of intepreting a data element write. It expresses that a runnable set up to trigger on data reception on required data element $i.p.e$ should be marked as {\bf pending} whenever a write occurs on an element connected to $i.p.e$. An example may be as follows (assuming $\Tconnect {i.p.e} a$ and $\TdataReceivedEvent b a$:
$$
\begin{array}{c}
	\Trinst {i.r} c \Xs {\TrteWrite {p.e} v k} \,\Opar  \\
	{\Tdelem a \_ \_} \Opar {\Trunnable b t \_ n} \vspace{0.6ex} \\
	\red{\say{i.p.e}{\Twr v}} \\
	\Trinst {i.r} c \Xs {\Tap k \Tvoid} \,\Opar \\
	{\Tdelem a \Ttrue v} \Opar {\Trunnable b t \Tpending n}
\end{array}
$$
Transitions that actually create an instance of a pending runnable will be introduced in Section~\ref{sect:SpawnTerm}.

@@

Buffered interprocess communication (axioms (22) - (30)) is essentially a minor variant of the unbuffered mechanism, where the single-value store of a {\bf delem} term is replaced by a {\bf qelem} process holding a sequence of values. The buffer is consumed from the left (axiom (23)) and extended to the right (axiom (27)). When the buffer is empty, the special value (error code) {\bf nodata} is returned to receiving runnable instances (axiom (24)). When the buffer is at its maximum capacity, the communicated value is simply dropped (axiom (28)).

However, the AUTOSAR specification requires the send command to inform its caller whether all connected buffers were able to successfully store the communicated value or whether some buffers had it dropped. Such information is naturally expressed as an additional parameter $\Tvar{as}$ in the {\em snd} transition, listing the addresses of connected {\bf qelem} processes that have no spare capacity. The preconditions of axioms (27) and (28) ensure that $\Tvar{as}$ reflects the truth, and axioms (25) and (26) feed different error codes to the sender's continuation depending on whether $\Tvar{as}$ is empty or not. Precondition $\Tconnect {i.p.e} \Tvar{as}$ in axiom (26) should be read $\Tconnect {i.p.e} a$ for all $a \in \Tvar{as}$, and its purpose is to outrule irrelevant addresses in $\Tvar{as}$ causing bogus {\bf limit} results.

The following example illustrates a scenario where a sent value is only stored in a subset of the connected buffers. Assuming $\Tconnect {\Tep e p i} {a_1}$ and $\Tconnect {\Tep e p i} {a_2}$:
$$
\begin{array}{c}
	\Trinst {i.r} c \Xs {\TrteSend {p.e} v k} \,\Opar  \\
	{\Tqelem {a_1} 2 \Teps} \Opar {\Tqelem {a_2} 2 {\Tcons {v_1} {v_2}}} \vspace{0.6ex} \\
	\red{\say{i.p.e}{\Tsnd v {a_2}}} \\
	\Trinst {i.r} c \Xs {\Tap k \Tlimit} \,\Opar \\
	{\Tqelem {a_1} 2 v} \Opar {\Tqelem {a_2} 2 {\Tcons {v_1} {v_2}}}
\end{array}
$$

@@

The behavior of client-server communication in AUTOSAR depends on whether a client lists the required server port operation as a synchronous or an asynchronous callpoint (which are mutually exclusive model attributes). In the former case, the call command is treated as if it were immediately followed by a command for retrieving the server result (axiom (31)), whereas in the latter case, the call succeeds immediately and the result has to be retireved explicitly by the client's continuation code (axiom (32)). In both cases, however, an attempt to call the server before it is done processing a previous asynchronous call leads to immediate abortion with the error code {\bf limit} (axiom (35)).
